---
title: "BaseR"
author: "Rob Hayward"
date: "10 June 2018"
output: html_document
---
##Base R
This is a test of the series called *Base R* that begins [here](https://jozefhajnala.gitlab.io/r/r001-reshape/)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r base1}
gdi <- read.csv(
  stringsAsFactors = FALSE, 
  url("https://jozefhajnala.gitlab.io/r/post/data/ESA2010_GDI.csv")
)
head(gdi[, 1:7])
```
To re-shape from wide to long, 
```{r reshape}
gdi_long <- reshape(data = gdi, 
                    direction = 'long',
                    varying = 2:23) # columns that will be stacked into 1
head(gdi_long)
```
The way that the data is presented can be improved. 
```{r reshape2}
gdi_long_full <- reshape(data = gdi, 
                         direction = "long", 
                         varying = 2:23, 
                         idvar = "country", #what identifies the rows
                         v.names = "GDI", # what are the columns called
                         timevar = "year", # what is the time column called
                         times = 1995:2016 # values for timevar
                         )
head(gdi_long_full)
```
It is also possible to go the other way, 
```{r, wide}
gdi_wide <- reshape(data = gdi_long_full, 
                    direction = "wide", 
                    idvar = "country", 
                    timevar = "year", 
                    v.names = "GDI" 
                    )
head(gdi_wide[, 1:7], drop = FALSE)
```
###Advanced Re-shape
This is a more challenging example
```{r challenge}
more_not_pretty <- read.csv(
  stringsAsFactors = FALSE,
  file = "https://jozefhajnala.gitlab.io/r/post/data/ESA2010_not_pretty.csv"
)
head(more_not_pretty[,1:5], drop = FALSE)
```
Now the variables do not have easily recognisable names.  They have to be specified as a list of vectors with each element specificying the columns for one varying vector with ``v.names`` for each of the variables. 
```{r not_pretty}
more_not_pretty_long <- reshape(
  data = more_not_pretty,
  direction = "long", 
  varying = list(2:23,
                 24:45,
                 46:67),
  timevar = "year",
  times = 1995:2016,
  idvar = "country",
  v.names = c("ConspC",
              "AGDIpC",
              "GrossSaving"
              )
)
head(more_not_pretty_long)
```
Alternatively, 

```{r not_pretty_two}
more_pretty <- read.csv(
  stringsAsFactors = FALSE,
  file = "https://jozefhajnala.gitlab.io/r/post/data/ESA2010_not_pretty.csv")
head(more_pretty[, 1:5, drop = FALSE])
```

```{r not_pretty_three}
more_pretty_long <- reshape(data = more_pretty, 
                           direction = "long", 
                           varying = 2:67,
                           idvar = "country"
                           )
head(more_pretty_long)
```


### Nairve Bayes





