---
title: "DataFrame"
author: "Rob Hayward"
date: "7 November 2015"
output: 
  html_document: 
    theme: cosmo
---

From [Lionel Hertzog](http://datascienceplus.com/working-with-data-frame-in-r/)

##Transpose
```{r}
data(cars)
head(mtcars)
mtcarst <- t(mtcars)
# cut some columns to print nicely
head(mtcarst[,1:4])
```
##Merge
```{r}
data(cars)
df <- data.frame(Color = c(rep("red", 5), rep("grey", 10), rep("white", 10), 
                           rep("black", 5), rep("chrome", 2)))
df$Model <- rownames(mtcars)
mtcars$Model <- rownames(mtcars)
df_merge <- merge(mtcars, df)
# cut some columns to print nicely.
head(df_merge[,c(1, 2, 3, 4, 5, 6, 7, 11, 12, 13)])
```
Merge works here beause the first row names are the same. Otherwise, you would have to specify the column that is to be used for the merging. `by.x` gives the name of the column of the first dataframe to use and `by.y` is the name of the column of the second dataframe to use.  If there are more rows in the second dataframe, it is necessary to make `all.y` TRUE so that rows that do not git with the first dataframe are added and filled with NAs. In the following example `all.x` are not necessary. 

````{r}
colnames(df)[2] <- "Models"
df[33,] <- c("grey", "Citron")
df_merged <- merge(mtcars, df, by.x = "Model", by.y = "Models", all.x = TRUE,
                   all.y = TRUE)
df_merged[df_merged$Model == "Citron", ]
```
##Sorting
To sort (ascending) on the number of gears, 

```{r}
df_ordered <- df_merged[order(df_merged$gear), ]
df_merged$gear
df_ordered$gear
```
The function `ordered` returns a vector of indices that will order the dataframe. This can also be used with strings to order alphabetically. 

##Change column order
```{r}
df_merged <- df_merged[, c(1, 13, 2:12)]
head(df_merged[ ,1:10])
```
###Delete a column
```{r}
df_merged$Color <- NULL
head(df_merged[, c(1, 5:11)])
```
##Subsetting or indexing
Can use square braket or `subset` function
```{r}
new <- df_merged[df_merged$gear >= 4, ]
new2 <- subset(df_merged, gear >= 4)
new$gear
new2$gear
```
Can also use the function `%in%` to get the intersection between a vector and a data.frame. 
```{r}
n <- 8
new <- df_merged[df_merged$cyl %in% n, ]
head(new)
```
There may be multiple arguments used to extract components of the dataframe
```{r}
new <- subset(df_merged, gear == 4 & cyl != 8)
head(new)
```