---
title: "Learning R"
author: "Rob Hayward"
date: "Thursday, August 28, 2014"
output: html_document
---
# Some notes on \lstinline{dplyr}
This comes from [Roger Penn}(https://www.youtube.com/watch?v=aywFompr1F4&list=TLmZovpghOuEo)

# Simple Simulation
This comes from [>>R](http://www.r-bloggers.com/simple-probs/).  It is a very simple example of simultion that can be expnded upon.  There are a number of steps: 
* Collect the latest polls
* Set the estimated percentage
* Set the standard deviation
* Create a function to simulate the election
* Simulate
* Calculte the number of times tht Dillema wins

```{r}
dimela
```
# The apply family
This comes from [Rolling your Rs](http://rollingyours.wordpress.com/2014/05/15/the-apply-command-101/).  ```apply``` is applied to a matrix or dataframe.  It returns a matrix or vector.  To take the mean of a column
```{r}
set.seed(1)
mymat <- matrix(round(rnorm(16, 10), 2), 4, 4)
apply(mymat, 2, mean)
```
The number 2 indicates that it is run across columns.  It would be 1 for rows. Addditional arguments can be passed to the function.  For example, ```na.rm ``` or ```trim```.   
```{r}
apply(mymat, 2, mean, na.rm = T, trim = 0.5)
````
Functions can also be passed anonymously.  
```{r}
apply(mymat, 2, function(x) x/sum(x))
```
Final example, 
```{r}
url <- "http://steviep42.bitbucket.org/data/su_et_al_2002.txt"
mydata <- read.csv(url, header = T)
str(mydata)
apply(mydata, 2, function(x) c(mu = mean(x), sd = sd(x)))
```


# lapply command 101
This comes from [Rolling your Rs](http://rollingyours.wordpress.com/2014/10/20/the-lapply-command-101/)
This is a very important way of dealing with lists. This is the way to hold hetrogenous data. Create a list for family attributed. 
```{r}
family1 <- list(name = "Jones", numofchild = 2, ages = c(5, 7), 
                measles = c("Y", "N"))
```
The ```lapply``` function can be applied to this list. The following loops through the list to find the class of each element.

```{r}
str(lapply(family1, typeof))
```
To show how to pass an anonymous function to ```lapply```, 
```{r} 
str(lapply(family1, function(x) typeof(x)))
```
The function can also be defined in advance.
```{r} 
simplefunc <- function(x){
  mytype <- typeof(x)
  return(mytype)
}
lapply(family1, simplefunc)
```
It is possible to create a list of families
```{r}
family2 <- list(name="Espinoza",numofchild=4,ages=c(5,7,9,11),measles=c("Y","N","Y","Y"))
family3 <- list(name="Ginsberg",numofchild=3,ages=c(9,13,18),measles=c("Y","N","Y"))
family4 <- list(name="Souza",numofchild=5,ages=c(3,5,7,9,11),measles=c("N","Y","Y","Y","N"))
allfams <- list(f1=family1,f2=family2,f3=family3,f4=family4)
lapply(allfams, function(x) mean(x$ages))
```
The ```lapply``` function can be used effectivley with the ```split``` function.  ```split``` will return a list of dataframes.  For example, it is possible to split the mcars dataframe into cylinder types. 
```{r}
unique(mtcars$cyl)
mydfs <- split(mtcars, mtcars$cyl)
str(mydfs, max.level = 1)
lapply(mydfs, function(x) mean(x$mpg))
```
Or more advanced
```{r}
tapply(mtcars$mpg, mtcars$cyl, mean)
```


#Tidy and dplyr from HW
This comes from [R-Studio video.  ](http://pages.rstudio.net/Webinar-Series-Recording-Essential-Tools-for-R.html?mkt_tok=3RkMMJWWfF9wsRonuqzMZKXonjHpfsX56uovXKO2lMI%2F0ER3fOvrPUfGjI4ATMFlI%2BSLDwEYGJlv6SgFTbDMMa541LgKWxc%3D)
Not sure how long this will last. The first part is about the package ```tidyr```.  The three main commands are **gather**, **spread** and **separate**.  There is a 30-page document about *tidy data*.   There can be a link between ```tidyr``` and ```dplyr``` and ```ggvis```.  
 
The process of data analysis has three stages:  think it, describe it and do it. The commands for dplyr

* filter:  keep rows matching criteria
* select: pick columns by name
* arrange: reorder rows
* mutate: add new vaiables
* summarise: reduce varibles to vaules

This reduceds the **thinking** part of the data analysis. 


For the presentation of data use ```ggvis```.  This can flow from *The Grammer of Graphics*. There are links between ```ggplot2``` and ```ggvis```.  All the data are available in [github.](https://github.com/rstudio/webinars)  
#'dplyr intro from Rstudio
This comes from [R-Studio.](http://blog.rstudio.org/2014/01/17/introducing-dplyr/). The aim is to draw the list of baseball players with the most games. 
```{r, message=FALSE, warning=FALSE}
library(Lahman)
library(dplyr)
players <- group_by(Batting, playerID)
games <- summarise(players, total = sum(G))
head(arrange(games, desc(total)), 5)
```
Operations can be chained together with the ```%.%``` operator. For example, 
```{r}
Batting %.%
  group_by(playerID) %.%
  summarise(total = sum(G)) %.%
  arrange(desc(total))%.%
  head(5)
```
From the introductory vignette that can be accessed by ```vignette("introduction", package = "dplyr")```




##Video overview
This is the video from [R-videos](http://practicalrvideos.blogspot.co.uk/2014/08/dplyr-gamechanger-for-data-manipulation.html)

Load the Titanic data.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
train = read.csv("../../Data/train.csv")
head(train)
```
key functions in dplyr

* *select:* select columns from a dataframe
* *filter:* select rows from a data frame based upon criteria
* *group_by:* group by a factor variable
* *summarize:* allows you to do summary stats based upon the grouped variable
* *arrange:*  a better way to order the data set

### Selecting columns and rows of a dataframe 
Selecting columns
```{r}
VariablesThatICareAbout = select(train, Survived, Pclass, Sex, Age, SibSp, Parch, Fare) 
head(VariablesThatICareAbout)
```
Everything except for PassengerId:
```{r}
VariablesThatICareAbout = select(train, -PassengerId) 
head(VariablesThatICareAbout)
```
Columns Survived to Age
```{r}
VariablesThatICareAbout = select(train, Survived:Age) 
head(VariablesThatICareAbout)
```

Selecting rows: only first class passengers
```{r}
FirstClass = filter(train, Pclass == 1)
head(FirstClass)
```
Only first class male passengers
```{r}
FirstClass = filter(train, Pclass == 1 & Sex == "male")
head(FirstClass)
```

Ordering the rows by fare and then class.  To switch round the put ```-Fare```
```{r}
train = arrange(train, Fare, Pclass)
head(train)
```


###Multiple commands in a single step 
The main value is the ability to pipe the output from one command into another
use ```%>%``` to pipe one command into another.  This means that the ```select``` statement will be piped into the ```group-by``` statement and that will be piped into the ```summarize``` statement. 


Average fare by passenger class
```{r}
select(train, Pclass, Fare, Survived) %>%
  group_by(Pclass) %>% #don't have to specify the dataset a second time
  summarize(AvgFare = mean(Fare), ProbSurvived = mean(Survived), N = length(Fare))
```

Just change the ```group-by``` statement to look at the survival by sex.
```{r}
select(train, Sex, Survived) %>%
  group_by(Sex) %>%
  summarize(ProbSurvived = mean(Survived))
```
Both categories at the same time!
(group by two variables)
```{r}
A = select(train, Pclass, Sex, Fare, Survived) %>%
  group_by(Pclass, Sex) %>% #don't have to specify the dataset a second time
  summarize(AvgFare = mean(Fare), ProbSurvived = mean(Survived), N = length(Fare))
A
```

Variable creation 
Use the mutate function
```{r}
train = mutate(train
               , AgeByFare = Age * Fare
               , Age2 = Age^2)
#allows you to create variables in one step 
#without specifying the data frame a bunch of times
train$Age2 = train$Age^2
```

  ### ggplot teaser 
There will be another tutorial on ggplot
Just wanted to show this to motivate using ggplot!
dplyr makes it easier to produce grouped plots in ggplot
```{r}
ggplot(A, aes(x = Pclass, y = ProbSurvived)) + 
  geom_point(aes(size = N)) + 
  geom_line(aes(by = Sex, color = Sex))
```
Survival probability unaffected for male 2 or 3 class, 
but big difference for females in 2nd and 3rd class
